Supplementary Note: Ancestry inference and grouping from principal component analysis of genetic data

### 2021-01-12 paper/supp-note-ancestry-grouping.tex
make as supp note

\usepackage{xr}
\externaldocument{paper-ancestry-matching-supp}

@@

\title{Ancestry inference and grouping\\from principal component analysis of genetic data
}
\author{Florian Priv\'e$^{\text{1,}*}$}

@@

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

\noindent Contact:
\begin{itemize}
\item \url{florian.prive.21@gmail.com}
\end{itemize}

\vspace*{5em}

\abstract{
	Here we propose a simple, robust and effective method for global ancestry inference and grouping from Principal Component Analysis (PCA) of genetic data.
	The proposed approach is particularly useful for methods that need to be applied in homogeneous samples.
	First, we show that Euclidean distances in the PCA space are proportional to $F_{ST}$ between populations.
	Then, we show how to use this PCA-based distance to infer ancestry in the UK Biobank and the POPRES datasets.
	We propose two solutions, either relying on projection of PCs to reference populations such as from the 1000 Genomes Project, or by directly using the internal data.
	Finally, we conclude that our method and the community would benefit from having an easy access to a reference dataset with an even better coverage of the worldwide genetic diversity than the 1000 Genomes Project.
}

@@

\newpage

\subsection*{Introduction}

@@

\vspace*{2em}
$^\dag$ Further defined in supplementary section ``Definitions $\dag$ and methods''.

@@

One could do the same using the countries of birth instead of the self-reported ancestries.
Again, the country of birth may sometime not reflect the ancestral origin. Therefore, we first compute the robust centers (geometric medians) of all countries with at least 300 individuals. Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset of centers with good coverage of the overall dissimilarities (Figure \ref{fig:heatmap-country}).
Based on the previous clustering and the available sample sizes, we chosed to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
Only 2.8\% of all individuals could not be matched (Table \ref{tab:country-groups}).
The resulting clusters are presented in the PCA space in figure \ref{fig:grouping2}.
Note that these clusters probably include individuals from nearby countries as well.

@@

\section*{Software and code availability}

The newest version of R package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
The code used in this paper is available at \url{https://github.com/privefl/paper-ancestry-matching/tree/master/code}.

\section*{Acknowledgements}

Authors thank Alex Diaz-Papkovich, Clive Hoggart and others for their useful feedback.
This research has been conducted using the UK Biobank Resource under Application Number 41181.

\section*{Funding}

F.P.\ is supported by the Danish National Research Foundation (Niels Bohr Professorship to John McGrath).

\section*{Declaration of Interests}

The authors declare no competing interests.

@@

%\newpage

@@

>>>>

@@

\title{Supplementary Note: Ancestry inference and grouping from principal component analysis of genetic data
}

@@

@@

@@

\vspace*{2em}
$^\dag$ Further defined in section ``Definitions $\dag$ and methods''.

@@

{\bf One could do the same using the countries of birth instead of the self-reported ancestries, which we use in the main text.}
Again, the country of birth may sometime not reflect the ancestral origin. Therefore, we first compute the robust centers (geometric medians) of all countries with at least 300 individuals. Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset of centers with good coverage of the overall dissimilarities (Figure \ref{fig:heatmap-country}).
Based on the previous clustering and the available sample sizes, we chosed to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
Only 2.8\% of all individuals could not be matched (Table \ref{tab:country-groups}).
The resulting clusters are presented in the PCA space in figure \ref{fig:grouping2}.
Note that these clusters probably include individuals from nearby countries as well.

\input{../tables/country-groups}

\begin{figure}[p]
	\centerline{\includegraphics[width=0.9\textwidth]{UKBB-matched-country}}
	\caption{The first eight PC scores computed from the UK Biobank (Field 22009) colored by the homogeneous ancestry group we infer for these individuals. \label{fig:grouping2}}
\end{figure}

@@

\subsection*{Definitions $\dag$ and methods}

Note that the code used in this supplementary note is available at \url{https://github.com/privefl/paper-ancestry-matching/tree/master/code}.

\begin{itemize}

	\item The {\bf 1000 Genomes Project (1000G)} data is composed of approximately 100 individuals for each of 26 populations worldwide (described at \url{https://www.internationalgenome.org/category/population/}), including 7 African (AFR), 5 East Asian (EAS), 5 South Asian (SAS), 5 European (EUR) and 4 admixed American (AMR) populations.
	Here we used the transformed data in PLINK format provided in \cite{prive2020efficient}.

	\item The $\boldsymbol{F_{ST}}$ measures the relative amount of genetic variance between populations compared to the total genetic variance within these populations \cite[]{wright1965interpretation}.
	We use the weighted average formula proposed in \cite{weir1984estimating}, which we now implement in our package bigsnpr \cite[]{prive2017efficient}.

	\item The {\bf Principal Component (PC) scores} are defined as $U \Delta$, where $U \Delta V^T$ is the singular value decomposition of the (scaled) genotype matrix \cite[]{prive2020efficient}. They are usually truncated, e.g.\ corresponding to the first 20 principal dimensions only.

	\item The {\bf Bhattacharyya distance} between two multivariate normal distributions $\mathcal{N}(\boldsymbol\mu_1,\,\boldsymbol\Sigma_1)$ and $\mathcal{N}(\boldsymbol\mu_2,\,\boldsymbol\Sigma_2)$ is defined as
	$D_B={1\over 8}(\boldsymbol\mu_2-\boldsymbol\mu_1)^T \boldsymbol\Sigma^{-1}(\boldsymbol\mu_2-\boldsymbol\mu_1)+{1\over 2}\log \,\left({|\boldsymbol\Sigma| \over \sqrt{|\boldsymbol\Sigma_1| \, |\boldsymbol\Sigma_2|} }\right)$,
	where $\boldsymbol\Sigma={\boldsymbol\Sigma_1+\boldsymbol\Sigma_2 \over 2}$ and $|M|$ is the absolute value of the determinant of matrix $M$ \cite[]{bhattacharyya1943measure,fukunaga1990introduction}. 
	The mean and covariance parameters for each population are computed using the robust location and covariance parameters as proposed in \cite{prive2020efficient}.

	\item The {\bf geometric median} of points is the point that minimizes the sum of all Euclidean distances to these points. We now implement this as function \texttt{geometric\_median} in our R package bigutilsr.

	\item The {\bf UK Biobank} is a large cohort of half a million individuals from the UK, for which we have access to both genotypes and multiple phenotypes (\url{https://www.ukbiobank.ac.uk/}).
	We apply some quality control filters to the genotyped data; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
	This results in 488,371 individuals and 504,139 genetic variants.

\end{itemize}

@@

@@

\section*{Additional Figures and Tables}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{equation}{0}
\renewcommand{\thesection}{S\arabic{section}}
\setcounter{section}{1}

\input{../tables/fst-values-AFR}
\input{../tables/fst-values-AMR}
\input{../tables/fst-values-EAS}
\input{../tables/fst-values-EUR}
\input{../tables/fst-values-SAS}

\FloatBarrier

\begin{figure}[h]
	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-Fst-1000G}}
	\caption{Heatmap with clustering based on the $F_{ST}$ between pairs of the 26 1000G populations. Corresponding values are reported in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}. \label{fig:heatmap0}}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-bhattacharyya-1000G}}
	\caption{Heatmap with clustering based on the Bhattacharyya distances between pairs of the 26 1000G populations. \label{fig:heatmap1}}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.7\textwidth]{compare-Bhattacharyya-to-Fst}}
	\caption{Comparing $F_{ST}$ to the Bhattacharyya distance on the PCA space between pairs of the 26 1000G populations. \label{fig:compare-dist1}}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-centers-1000G}}
	\caption{Heatmap with clustering based on the Euclidean distances between centers of pairs of the 26 1000G populations. \label{fig:heatmap2}}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{compare-Euclidean-to-Fst2}}
	\caption{Comparing $F_{ST}$ to the squared Euclidean distances on the PCA space between centers of pairs of the 26 1000G populations. Distances are computed using different numbers of Principal Components (PCs). \label{fig:compare-distK}}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.8\textwidth]{heatmap-closest-1000G}}
	\caption{Heatmap with clustering based on the shortest distances between individuals in pairs of the 26 1000G populations. \label{fig:heatmap3}}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.7\textwidth]{compare-closest-to-Fst}}
	\caption{Comparing $F_{ST}$ to the shortest distances between individuals in pairs of the 26 1000G populations. \label{fig:compare-dist3}}
\end{figure}

\FloatBarrier

\begin{figure}[h]
	\centerline{\includegraphics[width=0.9\textwidth]{proj-UKBB-to-1000G}}
	\caption{First 18 PC scores of the 1000G data (in black), onto which the UK Biobank data has been projected (in red). \label{fig:proj-UKBB}}
\end{figure}

\FloatBarrier

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{UKBB-mismatch}}
	\caption{PC scores (computed in the UK Biobank) colored by self-reported ancestry. On the left, these are 50,000 random individuals. On the right, these are the 47 individuals with some discrepancy between their self-reported-ancestry and our ancestry estimation (see table \ref{tab:infer-UKBB-superpop}). \label{fig:mismatch}}
\end{figure}

\input{../tables/matched-pop-KNN}
\input{../tables/infer-UKBB-pop}
\input{../tables/POPRES-matching}

\FloatBarrier

\begin{figure}[h]
	\centerline{\includegraphics[width=0.9\textwidth]{heatmap-country-UKBB}}
	\caption{Heatmap with clustering based on the distances in the PCA space between centers of pairs of the countries of birth in the UK Biobank. \label{fig:heatmap-country}}
\end{figure}

\input{../tables/ancestry-groups}

\FloatBarrier

\begin{figure}[h]
	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-overall-center}}
	\caption{Histogram of (log) squared distances from the UK Biobank PC scores to the geometric median of the all UKBB individuals. Here we use a threshold at 7, based on visual inspection. Alternatively, a more stringent threshold at 6 could also be used. \label{fig:hist3}}
\end{figure}
